---
title: "Relation Ontology Usage"
author: "Nicolas Matentzoglu"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    html_document: default 
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE}
library(data.table)
library(stringr)
library(plyr)
library(ggplot2)
library(DT)
library(knitr)
```


```{r settings, echo=FALSE}
ws = "rosurvey_obo.RData"
ro_geo = "rosurvey_corpus_obo.csv"


p_corpus="obo"
p_imports="merged"

fontsize=2

rbox_types<-c("SubPropertyChainOf", "SubObjectPropertyOf", "ObjectPropertyDomain","EquivalentObjectProperties","ObjectPropertyRange", "InverseObjectProperties", "TransitiveObjectProperty", "ReflexiveObjectProperty", "AsymmetricObjectProperty", "IrrefexiveObjectProperty", "InverseFunctionalObjectProperty", "FunctionalObjectProperty", "SymmetricObjectProperty", "DisjointObjectProperties")

rule_types<-c("Rule")
abox_types<-c("ClassAssertion", "ObjectPropertyAssertion")
tbox_types<-c("SubClassOf", "EquivalentClasses","DisjointClasses")

characteristics_axiom_types<-c("TransitiveObjectProperty", "ReflexiveObjectProperty", "AsymmetricObjectProperty", "IrrefexiveObjectProperty", "InverseFunctionalObjectProperty", "FunctionalObjectProperty", "SymmetricObjectProperty")

options(scipen = 1000)
opts_chunk$set(fig.width=8, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE)
```

```{r functions, echo=FALSE}
pc<-function(x,y,digits=2) {
  return(round(100*(x/y),digits))
}

theme_ro<-function(p) {
  cls<-c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
  p + theme_bw()+theme(axis.text.x = element_text(size=12),axis.text.y = element_text(size=12),axis.title = element_text(size=12)) + scale_fill_brewer(palette = "Paired") + scale_colour_brewer(palette = "Paired")
 # +  scale_color_manual(values=cls) +   scale_fill_manual(values=cls)
}

kt<-function(x) {
  return(kable(x,row.names = FALSE,digits = c(2)))
}
```

```{r load, echo=FALSE, cache=TRUE}
load(ws)
df_geo<-read.csv(file = ro_geo,stringsAsFactors = FALSE)
ro_labels<-unique(df_p[,c("p","label")])
df_p<-df_p[corpus==p_corpus,]
df_ax<-df_ax[corpus==p_corpus,]
```

```{r pre_geodata, echo=FALSE}
df_geo$lon<-ifelse(!is.na(df_geo$geolocation),as.character(as.data.frame(str_split_fixed(df_geo$geolocation, ",", 2))[[1]]),NA)
df_geo$lat<-ifelse(!is.na(df_geo$geolocation),as.character(as.data.frame(str_split_fixed(df_geo$geolocation, ",", 2))[[2]]),NA)
df_geo$lon<-as.numeric(df_geo$lon)
df_geo$lat<-as.numeric(df_geo$lat)
d<-df_geo[,c("geolocation","lon","lat","o")]
df_ct<-count(d[,c("geolocation","lon","lat")])
```

```{r pre_obocorpus, echo=FALSE}
d_obo<-ct_download_sucess[ct_download_sucess$corpus=="obo",]
all<-(df_corpus$o)
notdownloaded<-(df_corpus[!df_corpus$download_success,]$o)
downloadnotmerged<-(df_corpus[df_corpus$download_success&!df_corpus$merged_success,]$o)
mergednotcollected<-(df_corpus[df_corpus$merged_success&!df_corpus$collectdata_success,]$o)
collectedbutexcluded<-(df_corpus[df_corpus$collectdata_success&df_corpus$excluded,]$o)
ct_ont_all<-length(all)
ct_ont_download_fail<-length(notdownloaded)
ct_ont_merge_fail<-length(downloadnotmerged)
ct_ont_collect_fail<-length(mergednotcollected)
ct_ont_excluded<-length(collectedbutexcluded)

st_ont_all<-paste(all,collapse=" ")
st_ont_download_fail<-paste(notdownloaded,collapse=" ")
st_ont_merge_fail<-paste(downloadnotmerged,collapse=" ")
st_ont_collect_fail<-paste(mergednotcollected,collapse=" ")
st_ont_excluded<-paste(collectedbutexcluded,collapse=" ")

ct_corp_obo<-ct_ont_all-ct_ont_download_fail-ct_ont_merge_fail-ct_ont_collect_fail-ct_ont_excluded
```

# The OBO Corpus
In the following we will describe the OBO corpus in general. The OBO corpus contains `r ct_ont_all` ontologies, 

* `r ct_ont_download_fail` of which failed to download: `r st_ont_download_fail`, 
* `r ct_ont_merge_fail` of which failed to be parsed by ROBOT (or were empty): `r st_ont_merge_fail`
* `r ct_ont_collect_fail` of which failed to be processed by the data collection pipeline (timeout 60min): `r st_ont_collect_fail`
* `r ct_ont_excluded` of which failed were manually excluded: `r st_ont_excluded` 

resulting in a total of `r ct_corp_obo` valid ontologies for the following analysis. Full breakdown:

```{r processingbd, echo=FALSE}
kt(d_obo)
```

OBO ontologies are maintained by groups and individuals all around the world. The following map gives an indication of where OBO ontologies are developed, albeit a weak one: 

1. Every ontology gets only *one geolocation* - while, of course, every ontology has multiple contributors throughout the world, which makes this a fairly extreme simplification.
2. The geolocation is determined by the single point of contact given by the OBO foundry metadata. If a point of contact was was present in the OBO foundry metadata, the geolocation was determined using their email address, or, if no email or a personal (non-institution) email was given, the geolocation was traced down through the repository link provided, and sometimes manually with google search. 

Therefore, the following graphic should be read with a grain of salt: (1) A large number of contributors may be missing from the plot and (2) the current affiliation of the main contact may be a very weak way to establish an ontologies geolocation. 


```{r geodata}
map <- map_data("world")
ggplot() + coord_quickmap(xlim = c(-125,30), ylim=c(-35,60)) + geom_polygon(data=map,aes(long, lat, group=group))+ geom_point(aes(x=df_ct$lat, y=df_ct$lon,colour=df_ct$freq),alpha=0.5, size=df_ct$freq)  + scale_colour_gradient(low = "orange", high="red") +xlab("") + ylab("") + labs(colour="Ontologies") + theme(axis.text.y = element_blank(),axis.ticks.y = element_blank(),axis.ticks.x = element_blank(),axis.text.x = element_blank(),legend.position = c(0.1, 0.2)) + ggtitle("Number of ontologies contributed to the OBO foundry (location of main contact)")
```



```{r prepro_obo, echo=FALSE}
# Analysis of OBO snapshot
# IN df_ax, df_p

df_ax$ax_cat<-ifelse((df_ax$axiomtype %in% tbox_types),"tbox",ifelse((df_ax$axiomtype %in% abox_types),"abox",ifelse((df_ax$axiomtype %in% rule_types),"rules",ifelse((df_ax$axiomtype %in% rbox_types),"rbox","other"))))
df_ax$tboxabox<-ifelse((df_ax$axiomtype %in% tbox_types)|(df_ax$axiomtype %in% abox_types),"tboxabox","other")
da<-df_ax
da$all="all"
dm<-df_ax[df_ax$ftype=="merged",]
dni<-df_ax[df_ax$ftype=="noimports",]
dm_noro_tbox_abox<-dm[!(dm$axiomtype %in% rbox_types),]
dm_noro_tbox_abox_ni<-dni[!(dni$axiomtype %in% rbox_types),]
dm_noro_tbox_abox$corpussize <- ct_corp_obo
dm_noro_tbox_abox_ni$corpussize <- ct_corp_obo
dsig_m<-df_p[df_p$ftype=="merged",c("p","o")]
dsig_ni<-df_p[df_p$ftype=="noimports",c("p","o")]

ct_allaxiomdata$ax_cat<-ifelse((ct_allaxiomdata$axiomtype %in% tbox_types),"tbox",ifelse((ct_allaxiomdata$axiomtype %in% abox_types),"abox",ifelse((ct_allaxiomdata$axiomtype %in% rule_types),"rules",ifelse((ct_allaxiomdata$axiomtype %in% rbox_types),"rbox","other_cat"))))
ct_allaxiomdata$tboxabox<-ifelse((ct_allaxiomdata$axiomtype %in% tbox_types)|(ct_allaxiomdata$axiomtype %in% abox_types),"tboxabox","other")
daa<-ct_allaxiomdata
daa$all="all"

ct_unique_o_using_rel<-length(unique(dm_noro_tbox_abox$o))
ct_unique_rel_tbox_abox<-length(unique(dm_noro_tbox_abox$p))
ct_unique_o_using_rel_ni<-length(unique(dm_noro_tbox_abox_ni$o))
ct_unique_rel_tbox_abox_ni<-length(unique(dm_noro_tbox_abox_ni$p))
# OUT dm, dni, dm_noro_tbox_abox, dm_noro_tbox_abox_ni, corpus, dsig_m, dsig_ni, ct_unique_o_using_rel, ct_unique_rel_tbox_abox, ct_unique_o_using_rel_ni, ct_unique_rel_tbox_abox_ni, 
```

# How important is the Relation Ontology (RO) for OBO foundry ontologies? 
We say that an axiom ax is RO-dependent if the axiom contains one or more RO-relations. An ontology is RO-dependent if it contains at least one RO-dependent axiom. We say that an axiom is p-dependent if it contains a particular object property p. In the following, we will quantify the importance of RO for OBO ontologies by (1) counting the number of RO-dependent ontologies and (2) determining the prevalence of RO-dependent axioms.

```{r breakdown_o, echo=FALSE, include=FALSE}
# OBO ontologies and RO relation usage. Ignored now in favour of axiom-based counting
#IN: dsig_m, dsig_ni
dsig_m_unique<-unique(dsig_m[,c("p","o")])
dsig_ni_unique<-unique(dsig_ni[,c("p","o")])
dsig_m_o_ct<-count(dsig_m_unique[,c("o")])
names(dsig_m_o_ct)<-c("o","ct_sig_m")
dsig_ni_o_ct<-count(dsig_ni_unique[,c("o")])
names(dsig_ni_o_ct)<-c("o","ct_sig_ni")
df_o<-merge(dsig_ni_o_ct,dsig_m_o_ct,by="o",all=TRUE)
df_o[is.na(df_o)] <- 0
#write.csv(file = "ro_survey_onts.csv",df_o)
names(df_o)<-c("Ontology","RO rels (no imports)","RO rels (merged imports)")
kt(df_o)
dsig_m_unique<-dsig_ni_unique<-dsig_m_o_ct<-dsig_ni_o_ct<-NULL
# OUT: df_o
```

## Analysis of RO-dependent ontologies

```{r rodeo, echo=FALSE}
df_rodeo<-unique(da[,.(o,ax_cat,ftype)])
ct_rodeo_axcat<-count(df_rodeo[,.(ax_cat,ftype)])
names(ct_rodeo_axcat)<-c("category","ftype","freq")

df_rodeo2<-unique(da[,.(o,tboxabox,ftype)])
ct_rodeo_tboxrbox<-count(df_rodeo2[,.(tboxabox,ftype)])
names(ct_rodeo_tboxrbox)<-c("category","ftype","freq")

df_rodeo3<-unique(da[,.(o,all,ftype)])
ct_rodeo_all<-count(df_rodeo3[,.(all,ftype)])
names(ct_rodeo_all)<-c("category","ftype","freq")

x<-merge(ct_rodeo_axcat,ct_rodeo_tboxrbox,by=c("category","ftype","freq"),all=TRUE)
x<-merge(x,ct_rodeo_all,by=c("category","ftype","freq"),all=TRUE)
x$pc<-pc(x$freq,ct_corp_obo,1)
d_rodeo<-x[x$category!="tboxabox",]
```

```{r rodeo_p, echo=FALSE}
p<-ggplot(d_rodeo,aes(x=reorder(category,freq),y=pc,fill=ftype,group=ftype)) + geom_bar(stat="identity",position="dodge") + geom_text(aes(label=paste("",freq," (",pc,"%)",sep="")), hjust=-0.1,vjust=0.5, color="black",position = position_dodge(width = 1),size=3.5) + scale_x_discrete(limits = c("abox","rules","tbox","rbox","all")) + xlab("") + ylab("% of OBO ontologies") + guides(fill=guide_legend(title="Imports")) + coord_flip()
theme_ro(p) + theme(legend.position = c(0.8, 0.2)) + ylim(0,100)
ggsave("rosurvey_rodeo.jpg",width = 4.5,height = 3.5)
```

## Analysis of RO-dependent axioms

```{r aggregate_all_axioms, echo=FALSE}

ct_allaxiomdata_axcat<-aggregate(daa$freq,by=list(daa$ftype,daa$corpus,daa$ax_cat),FUN=sum)
names(ct_allaxiomdata_axcat)<-c("ftype","corpus","category","freq")

ct_allaxiomdata_tboaxabox<-aggregate(daa$freq,by=list(daa$ftype,daa$corpus,daa$tboxabox),FUN=sum)
names(ct_allaxiomdata_tboaxabox)<-c("ftype","corpus","category","freq")

ct_allaxiomdata_all<-aggregate(daa$freq,by=list(daa$ftype,daa$corpus,daa$all),FUN=sum)
names(ct_allaxiomdata_all)<-c("ftype","corpus","category","freq")


x<-merge(ct_allaxiomdata_all,ct_allaxiomdata_tboaxabox,by=c("category","ftype","freq","corpus"),all=TRUE)
x<-merge(x,ct_allaxiomdata_axcat,by=c("category","ftype","freq","corpus"),all=TRUE)
x$corpus<-NULL
names(x)<-c("category","ftype","ax_ct")
d_roallax<-x[x$category!="tboxabox",]
```

```{r aggregate_ro_axioms, echo=FALSE}
df_rodeo<-unique(da[,.(o,ax_cat,axiomid,ftype)])
ct_rodeo_axcat<-count(df_rodeo[,.(ax_cat,ftype)])
names(ct_rodeo_axcat)<-c("category","ftype","freq")

df_rodeo2<-unique(da[,.(o,tboxabox,axiomid,ftype)])
ct_rodeo_tboxrbox<-count(df_rodeo2[,.(tboxabox,ftype)])
names(ct_rodeo_tboxrbox)<-c("category","ftype","freq")

df_rodeo3<-unique(da[,.(o,all,axiomid,ftype)])
ct_rodeo_all<-count(df_rodeo3[,.(all,ftype)])
names(ct_rodeo_all)<-c("category","ftype","freq") 

x<-merge(ct_rodeo_axcat,ct_rodeo_tboxrbox,by=c("category","ftype","freq"),all=TRUE)
x<-merge(x,ct_rodeo_all,by=c("category","ftype","freq"),all=TRUE)

x<-merge(x,d_roallax,by=c("category","ftype"))
x$pc<-pc(x$freq,x$ax_ct,1)
d_roax<-x[x$category!="other"&x$category!="tboxabox",]

```

```{r roaxplot, echo=FALSE}
p<-ggplot(d_roax,aes(x=reorder(category,freq),y=pc,fill=ftype,group=ftype)) + geom_bar(stat="identity",position="dodge") + geom_text(aes(label=paste("",freq,sep="")), hjust=-0.1,vjust=0.5, color="black",position = position_dodge(width = 1),size=3.5) + scale_x_discrete(limits = c("abox","rules","tbox","rbox","all")) + xlab("") + ylab("% of axioms across OBO-foundry") + guides(fill=guide_legend(title="Imports")) + coord_flip()
theme_ro(p) + theme(legend.position = c(0.85, 0.8)) + ylim(0,105)
ggsave("rosurvey_axiomcat.jpg",width = 4.5,height = 3.5)
```

# How important are the individual relations? Which RO relations are the most important?

```{r prep_plots, echo=FALSE}
#IN: dm_noro_tbox_abox, dm_noro_tbox_abox_ni
dm_noro_tbox_abox$corpussize<-ct_corp_obo
dm_noro_tbox_abox_ni$corpussize<-ct_corp_obo
dgg<-unique(dm_noro_tbox_abox[,c("p","label","o","corpussize")])
cdgg<-count(dgg[,c("p","label","corpussize")])
cdgg$pc<-round(pc(cdgg$freq,cdgg$corpussize),0)
cdgg$type<-"merged"
cdgg$corpussize<-NULL

dgg_ni<-unique(dm_noro_tbox_abox_ni[,c("p","label","o","corpussize")])
cdgg_ni<-count(dgg_ni[,c("p","label","corpussize")])
cdgg_ni$pc<-round(pc(cdgg_ni$freq,cdgg_ni$corpussize),0)
cdgg_ni$type<-"removed"
cdgg_ni$corpussize<-NULL

x<-rbind(cdgg,cdgg_ni)
dgg<-dgg_ni<-cdgg_ni<-cdgg<-NULL

x<-reshape(x, idvar = c("p","label"), timevar = "type", direction = "wide")
x$top20<-ifelse(x$freq.merged>25,TRUE,FALSE)
x<-melt(id.vars=c("p","label","top20"),x)
x$measure<-gsub(".merged","",gsub(".removed","",x$variable))
x$ftype<-gsub("freq.","",gsub("pc.","",x$variable))
x$variable<-NULL
x<-reshape(x, idvar = c("p","label","ftype","top20"), timevar = "measure", direction = "wide")
names(x)<-gsub("value.","",names(x))
x$freq<-ifelse(is.na(x$freq),0,x$freq)
x$pc<-ifelse(is.na(x$pc),0,x$pc)
x<-x[complete.cases(x), ]
#OUT: x (to be removed after plotting)
```


## Distribution of usage (nr of ontologies) for all RO relations 

Across TBOX and ABOX axioms, ignoring RBOX axioms!

```{r}
ggplot(x,aes(x=reorder(label,freq),y=freq,fill=ftype))+geom_bar(stat="identity",position = "dodge")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  xlab("RO Relation") + ylab("Number of ontologies used in") 
#ggsave("obo_all_mni.png", width = 15, height = 7, units = "cm")
```

## Relations used in more than 25 ontologies

Across TBOX and ABOX axioms, ignoring RBOX axioms!

```{r}
p<-ggplot(x[x$top20,],aes(x=reorder(label,freq),y=freq,fill=ftype))+geom_bar(stat="identity",position = "dodge")+ geom_text(aes(label=paste("",pc," %",sep="")), hjust=1, color="black",position = position_dodge(width = 1)) +xlab("RO Relation") + ylab("Number of ontologies used in") + coord_flip() 
theme_ro(p)
#ggsave("obo_20_plus_mni.png", width = 12, height = 12, units = "cm")
```

## Full breakdown of all relations with use in ontologies
```{r, echo=FALSE}
xl<-x
xl$top20<-NULL
xl$ftype<-NULL
xl<-xl[order(-xl$freq),]
names(xl)<-c("RO relation","Label","Nr. ontologies","% ontologies")
kt(xl)
```


```{r, echo=FALSE, include=FALSE}
### Number of ontologies by axiom types
# The following tables shows how many ontologies import or define a particular object property characteristic for ANY RO relation. For example, there might be 6 ontologies that define at least one RO relation as asymmetric.

d_char<-dm[dm$o!="ro"&(dm$axiomtype %in% characteristics_axiom_types),]
d_char<-unique(d_char[,c("axiomtype","label","o")])
d_char_ct<-count(d_char[,c("label","axiomtype")])
d_char_p<-unique(d_char[,c("axiomtype","label")])
d_char_p_ct<-count(d_char_p[,c("axiomtype")])
names(d_char_p_ct)<-c("Property Characteristic","Nr. ontologies")
kt(d_char_p_ct)
```

## Full Breadown of usage of RO relations
in_sig_m:	Number of ontologies where RO relation appears in signature, imports merged

in_sig_ni:	Number of ontologies where RO relation appears in signature, imports removed

tboxabox_merged	Number of ontologies where RO relation appears in TBox or Abox axiom (not RBox), imports merged

tboxabox_noimports	Number of ontologies where RO relation appears in TBox or Abox axiom (not RBox), imports removed

all other:	Number of ontologies using the relation as part of a particular axiom type, imports MERGED

The table is large. Note that you can scroll horizontally!

```{r, echo=FALSE}
# P occurs in signature (merged)
d_sig_ct<-count(dsig_m[,c("p")])
names(d_sig_ct)<-c("p","in_sig_m")

# P occurs in signature (no imports)
d_sig_ni_ct<-count(dsig_ni[,c("p")])
names(d_sig_ni_ct)<-c("p","in_sig_ni")

# P usage TBOX ABOX (imports merged)
ro_cts_m<-unique(dm_noro_tbox_abox[,c("p","o")])
ro_cts_m<-count(ro_cts_m[,c("p")])
names(ro_cts_m)<-c("p","tboxabox_merged")

# P usage TBOX ABOX (imports removed)
ro_cts_ni<-unique(dm_noro_tbox_abox_ni[,c("p","o")])
ro_cts_ni<-count(ro_cts_ni[,c("p")])
names(ro_cts_ni)<-c("p","tboxabox_noimports")

# P usage broken down by axiom type (imports merged)
ro_ctaxs<-unique(dm[,c("p","o","axiomtype")])
ro_ctaxs<-count(ro_ctaxs[,c("p","axiomtype")])
ro_ctaxs<-reshape(ro_ctaxs, idvar = "p", timevar = "axiomtype", direction = "wide")
names(ro_ctaxs)<-gsub("freq.","",names(ro_ctaxs))

ro_all<-merge(ro_labels,ro_cts_m,by="p",all=TRUE)
ro_all<-merge(ro_all,ro_cts_ni,by="p",all=TRUE)
ro_all<-merge(ro_all,d_sig_ct,by="p",all=TRUE)
ro_all<-merge(ro_all,d_sig_ni_ct,by="p",all=TRUE)
ro_all<-merge(ro_all,ro_ctaxs,by="p",all=TRUE)
ro_all[is.na(ro_all)] <- 0

ro_all<-ro_all[order(-ro_all$tboxabox_merged),]
kt(ro_all)
#write.csv(ro_all,file="ro.csv")

```

```{r cleanobo, echo=FALSE}
#x<-dm<-dni<-dm_noro_tbox_abox<-dm_noro_tbox_abox_ni<-corpus<-dsig_m<-dsig_ni<-ct_unique_o_using_rel<-ct_unique_rel_tbox_abox<-ct_unique_o_using_rel_ni<-ct_unique_rel_tbox_abox_ni<-NULL
```


```{r prepromain, echo=FALSE}
# IN: df_ax
ct_axt<-count(df_ax[,.(ftype,corpus,p,axiomtype,label)])
ct_axt<-ct_axt[order(-ct_axt$freq),]

df_axo<-unique(df_ax[,.(o,ftype,corpus,axiomtype,p)])
ct_axto<-count(df_axo[,.(ftype,corpus,p,axiomtype)])
ct_axto<-merge(ct_axto,df_ro,by.x = "p",by.y = "entity",all.x = TRUE)
ct_axto<-ct_axto[order(-ct_axto$freq),]

ct_p<-count(df_p[,.(ftype,corpus,p,label,corpussize)])
ct_p$pc<-pc(ct_p$freq,ct_p$corpussize)
#ct_p<-merge(ct_p,df_ro,by.x = "p",by.y = "entity",all = TRUE) //every relation is used at least once, else this would be necessary.
ct_p<-ct_p[order(-ct_p$freq),]

ct_o<-count(df_p[,.(ftype,corpus,o)])
ct_o<-ct_o[order(-ct_o$freq),]

df_po<-unique(df_p[,.(ftype,corpus,corpussize,o)])
ct_ro<-count(df_po[,.(ftype,corpus,corpussize)])
ct_ro$pc<-pc(ct_ro$freq,ct_ro$corpussize)

ct_ax<-count(df_ax[,.(ftype,corpus,p,label)])
names(ct_ax)<-c("ftype","corpus","p","label","axioms")

ct_p<-merge(ct_p,ct_ax,by=c("ftype","corpus","p","label"),all = TRUE)

```

# Other metrics (just dumped here)

## How many ontologies use RO?

This usage is based on whether an ontology contained a RO relation in its signature. This includes any axiom types (such as property characteristics), including declarations!
```{r, echo=FALSE}
x<- ct_ro
names(x)<-c("Imports","Corpus","Corpus size","Ontologies using RO","% Ontologies using RO")
kt(x)
```


## How many different RO relations does this ontology use?

Compared to the previous plot (very similar), this plot takes into account all axiom types, not just TBOX and ABOX!

```{r, echo=FALSE}
dgg<-df_ax
dgg<-merge(dgg,unique(df_p[,c("ftype","corpus","corpussize")]),by=c("ftype","corpus"),all.x = TRUE)
dgg<-unique(dgg[,c("label","o","corpussize")])
cdgg<-count(dgg[,c("label","corpussize")])
cdgg$pc<-pc(cdgg$freq,cdgg$corpussize)
#cdgg<-cdgg[cdgg$freq>=input$rel_ct_offset[1]&cdgg$freq<=input$rel_ct_offset[2],]
ggplot(cdgg[cdgg$freq>25,],aes(x=reorder(label,freq),y=freq))+geom_bar(stat="identity")+ geom_text(aes(label=paste(freq," (",pc," %)",sep="")), hjust=1.1, color="white") +xlab("RO Relation") + ylab("Number of ontologies used in") + coord_flip() + ggtitle("Top relations (used by more than 25 ontologies)")


asis_output("### The following table shows the number of distinct RO relations used across the selected axiom types")

asis_output("For example, the HP ontology might be using 45")

dgg<-unique(df_ax[,c("label","o")])
cdgg<-count(dgg[,c("o")])
#cdgg<-cdgg[cdgg$freq>=input$rel_ct_offset[1]&cdgg$freq<=input$rel_ct_offset[2],]
cdgg<-cdgg[order(-cdgg$freq),]
names(cdgg)<-c("Ontology","Number of RO relations used")
kt(cdgg)

```


```{r, echo=FALSE, include=FALSE}
## What kind of axiomtype is the RO relation used in? In how many axioms of that type is it used?
#IN: ct_axt (needs to be reactive!)
x<- ct_axt[,c("p","label","axiomtype","freq")]
names(x)<-c("RO relation","Label","Axiomtype","Number of axioms")
kt(x)
```

```{r, echo=FALSE, include=FALSE}
## What kind of axiomtype is the RO relation used in? In how many ontologies?
x<-ct_axto[,c("p","axiomtype","freq")]
names(x)<-c("RO relation","Axiomtype","Nr. of ontologies")
kt(x)

```

```{r, echo=FALSE}
asis_output("What kind of class expressions is the RO relation used in? In how many axioms of that type is it used?")
x<-unique(df_exp[,.(ftype,corpus,p,expressiontype,axiomid,o)])
x<-count(x[,.(ftype,corpus,p,expressiontype)])

x<-x[,c("p","expressiontype","freq")]
x<-merge(x,ro_labels,by=c("p"),all.x=TRUE)
x<-x[order(-x$freq),]
names(x)<-c("RO relation","Expression type","Nr. axioms","Label")
kt(x[,c("RO relation","Label","Expression type","Nr. axioms")])

asis_output("What kind of class expressions is the RO relation used in? In how many ontologies?")
x<-df_expo[,]
x<-count(x[,.(ftype,corpus,p,expressiontype)])
x<-x[,c("p","expressiontype","freq")]
x<-merge(x,ro_labels,by=c("p"),all.x=TRUE)
x<-x[order(-x$freq),]
names(x)<-c("RO relation","Expression type","Nr. ontologies","Label")
#kt(x[,c("RO relation","Label","Expression type","Nr. ontologies")])

```

# Domains and Ranges

## Asserted domains and ranges of RO relations
The asserted domains and ranges for a given RO relation in the ontology (i.e. ObjectPropertyDomain or ObjectPropertyRange axioms with that relation).
```{r domran, echo=FALSE}
x<-df_domain_range_asserted[,c("p","domain","range","o")]
x<-merge(x,ro_labels,by=c("p"),all.x=TRUE)
names(x)<-c("RO relation","Domain","Range","Ontology","Label")
kt(x[,c("RO relation","Label","Domain","Range","Ontology")])
```

## Inferred domains of OBO relations
The inferred domains for a given RO relation in the ontology. There are five places where domains can be inferred:
* ClassAssertion:
  * a:R some C --> domain is all types of a, range is C plus superclasses (types are all classes a is inferred to be a member of)
* DomainAxiom: 
  * any asserted domain and all their superclasses
* InNaryClassExpression: 
  * X and (R some D) and Y: Range is D (and superclasses), Domain is X and Y
* ObjectPropertyAssertion:     
  * <a,R,c> : Domain of R is types of a, Range is types of c 
* SubOrEquivalenceClass: 
  * A Sub R some C : Domain of R is A, Range is C
  * R some C Sub A : Domain of R is owl:Thing, Range is C
  * A eq R some C is treated as A sub R some C, R some C sub A 
  
```{r domaininf, echo=FALSE, include=FALSE}
x<-df_dom[,.(ftype,corpus,p,subject,expressiontype)]
x<-count(x)
x<-x[,c("p","subject","expressiontype","freq")]
x<-merge(x,ro_labels,by=c("p"),all.x=TRUE)
x<-x[order(-x$freq),]
names(x)<-c("RO relation","Domain","Source expression","Nr. ontologies","Label")
#kt(x[,c("RO relation","Label","Domain","Source expression","Nr. ontologies")])
```

## Inferred ranges of OBO relations
The inferred ranges for a given RO relation in the ontology. There are five places where ranges can be inferred:
* ClassAssertion:
  * a:R some C --> domain is all types of a, range is C plus superclasses (types are all classes a is inferred to be a member of)
* RangeAxiom: 
  * any asserted domain and all their superclasses
* InNaryClassExpression: 
  * X and (R some D) and Y: Range is D (and superclasses), Domain is X and Y
* ObjectPropertyAssertion:     
  * <a,R,c> : Domain of R is types of a, Range is types of c 
* SubOrEquivalenceClass: 
  * A Sub R some C : Domain of R is A, Range is C
  * R some C Sub A : Domain of R is owl:Thing, Range is C
  * A eq R some C is treated as A sub R some C, R some C sub A 

```{r rangeinf, echo=FALSE, include=FALSE}
x<-df_ran[,.(ftype,corpus,p,object,expressiontype)]
x<-count(x)
x<-x[,c("p","object","expressiontype","freq")]
x<-merge(x,ro_labels,by=c("p"),all.x=TRUE)
x<-x[order(-x$freq),]
names(x)<-c("RO relation","Range","Source expression","Nr. ontologies","Label")
#kt(x[,c("RO relation","Label","Range","Source expression","Nr. ontologies")])
```



```{r, echo=FALSE, include=FALSE}
# Full breakdown of relations used across ontologies
# A relation is 'used' by the ontology if it appears in its signature.
x<-unique(df_ax[,c("p","label","o","axiomid")])
x<-count(x[,c("p","label","o")])
names(x)<-c("RO relation", "Label","Ontology","Number of axioms")
#kt(x)
```

# Additional information
```{r, echo=FALSE}
asis_output("https://monarchinitiative.org/")
```
